\textbf{Effective application of machine learning to microbiome-based classification problems}

### 2019-10-16 submission/manuscript.tex
changes after table edtits from jenna

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to microbiome based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective application of machine learning to microbiome based
classification problems}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

2. Department of Electrical Engineering and Computer Science, University
or Michigan, Ann Arbor, MI 48109

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; unreported variation
between the predictive performance on different folds of
cross-validation; and unreported variation between cross-validation and
testing performances. Though, the microbiome field is making progress to
avoid some of these pitfalls including validating their models on
independent datasets (8, 19, 20) and introducing ways to better use ML
tools (21--24), more work is needed to further improve reproducibility
and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
(but not always) emphasize predictive performance. Nonetheless, it is
important to understand that the benefit of more complex, less
interpretable models may be minimal (29, 30). It is important for
researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and gradient boosted trees. We
compared their predictive performance, interpretability, and training
time. To demonstrate the performance of these modeling approaches and
our pipeline, we used data from a previously published study that sought
to classify individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (4). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals of which 52 had an SRN (54\%). The training
data were used to build and select the models and the test set was used
for evaluating the model.

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well (32). For regularized
logistic regression and SVM with a linear kernel, we used
L2-regularization to keep all potentially important features. For
comparison, we also trained an L1-regularized SVM with a linear kernel.
L1-regularization on microbiome data led to a sparser solution (i.e.,
forced many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including a decision tree, a random forest,
and gradient boosted trees (XGBoost) and an SVM with a non-linear
kernel.

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in previously developed ML packages available in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and should to be
optimized for each new ML task. For example, L1-regularized SVM with
linear kernel showed large variability between different regularization
strengths (C) and benefited from tuning {[}Figure S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
predictive performance of random forest model was higher than other ML
models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

@@

To evaluate the variation in the estimated performance, we calculated
the range of AUROC values for each model using 100 data-splits. The
range among the testing AUROC values within each model varied by 0.230
on average across the seven models. If we had only done a single split,
then there is a risk that we could have gotten lucky or unlucky in
estimating model performance. For instance, the lowest AUROC value of
the random forest model was 0.593 whereas the highest was 0.810. These
results showed that depending on the data-split, the testing performance
can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
data splits when estimating generalization performance.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
variation in cross-validation performance within both models at lower
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is limited to linear models or models
with prespecified interaction terms.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e., testing AUROC values) decreased when each OTU or group of OTUs
was randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680
which indicated high degree of collinearity in the dataset. Permutation
analysis allowed us to gauge the importance of each OTU in non-linear
models and partially account for collinearity by grouping correlated
OTUs to determine their impact as a group.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. We did not take advantage of
parallelization in training nor was the code optimized for speed. Still,
as expected, the training times increased with the complexity of the
model and the number of potential hyperparameter combinations. Also, the
linear models trained faster than non-linear models {[}Figures S1-S2;
Figure 5{]}.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other, however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (38). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 39, 40) but microbiome datasets often
suffer from having many features and small sample sizes, which result in
overfitting.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from:
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with normal colons
and those with screen relevant neoplasias (SRNs). The normal class
included patients with normal colons or non-advanced adenomas whereas
the SRN class included patients with advanced adenomas or carcinomas
(43). Patients with an adenoma greater than 1 cm, more than three
adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas. There were 172 patients with
normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
198 adenomas, 109 were identified as advanced adenomas. Together 261
patients were classified as normal and 229 patients were classified as
having a SRN.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to microbiome-based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective application of machine learning to microbiome-based
classification problems}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

2. Department of Electrical Engineering and Computer Science, University
of Michigan, Ann Arbor, MI 48109

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; unreported variation
between the predictive performance on different folds of
cross-validation; and unreported variation between cross-validation and
testing performances. Though the microbiome field is making progress to
avoid some of these pitfalls including validating their models on
independent datasets (8, 19, 20) and introducing ways to better use ML
tools (21--24), more work is needed to improve reproducibility further
and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
(but not always) emphasize predictive performance. Nonetheless, it is
essential to understand that the benefit of more complex, less
interpretable models may be minimal (29, 30). It is important for
researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest, and gradient boosted trees. We
compared their predictive performance, interpretability, and training
time. To demonstrate the performance of these modeling approaches and
our pipeline, we used data from a previously published study that sought
to classify individuals as having healthy colons or colonic lesions
based on the 16S rRNA gene sequences collected from fecal samples (4).
This dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had healthy colons (53\%) and 229 had a
screen relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals, of which 52 had an SRN (54\%). The training
data were used to build and select the models, and the test set was used
for evaluating the model.

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well (32). For regularized
logistic regression and SVM with a linear kernel, we used
L2-regularization to keep all potentially important features. For
comparison, we also trained an L1-regularized SVM with a linear kernel.
L1-regularization on microbiome data led to a sparser solution (i.e.,
forced many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including, a decision tree, a random forest,
and gradient boosted trees (XGBoost) and an SVM with a non-linear
kernel.

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in previously developed ML packages available in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and should to be
optimized for each new ML task. For example, L1-regularized SVM with
linear kernel showed large variability between different regularization
strengths (C) and benefited from tuning {[}Figure S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

@@

To evaluate the variation in the estimated performance, we calculated
the range of AUROC values for each model using 100 data-splits. The
range among the testing AUROC values within each model varied by 0.230
on average across the seven models. If we had only done a single split,
then there is a risk that we could have gotten lucky or unlucky in
estimating model performance. For instance, the lowest AUROC value of
the random forest model was 0.593, whereas the highest was 0.810. These
results showed that depending on the data-split, the testing performance
can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
data splits when estimating generalization performance.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
variation in cross-validation performance within both models at smaller
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is limited to linear models or models
with prespecified interaction terms.

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e., testing AUROC values) decreased when each OTU or group of OTUs
was randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680,
which indicated high degree of collinearity in the dataset. Permutation
analysis allowed us to gauge the importance of each OTU in non-linear
models and partially account for collinearity by grouping correlated
OTUs to determine their impact as a group.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g., OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. We did not take advantage of
parallelization in training, nor was the code optimized for speed.
Still, as expected, the training times increased with the complexity of
the model and the number of potential hyperparameter combinations. Also,
the linear models trained faster than non-linear models {[}Figures
S1-S2; Figure 5{]}.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (38). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 39, 40), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (43). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas. There were 172 patients with
normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
198 adenomas, 109 were identified as advanced adenomas. Together 261
patients were classified as healthy and 229 patients were classified as
having an SRN.
